ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-05-17 notebooks/gmsc.qmd
appendix
ufff

```{julia}
# Hyper:
_retrain = false

# Data:
test_size = 0.2
n_obs = Int(50000 / (1.0 - test_size))
counterfactual_data, test_data = train_test_split(load_gmsc(n_obs); test_size=test_size)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
output_dim = length(unique(labels))
```

@@

>>>>

```{julia}
# Hyper:
_retrain = true

# Data:
test_size = 0.2
counterfactual_data, test_data = train_test_split(load_gmsc(); test_size=test_size)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
output_dim = length(unique(labels))
```

@@

```{julia}
generator_params = DataFrame(
    Dict(
        :λ1 => λ₁,
        :λ2 => λ₂,
        :λ3 => λ₃,
        :opt => string(typeof(opt)),
        :eta => opt.eta,
        :dataname => "GMSC",
    )
)
CSV.write(joinpath(params_path, "generator/gmsc.csv"), generator_params)
```

### 2023-05-17 notebooks/mnist.qmd
ufff

```{julia}
function plot_mnist(
    factual::Int,target::Int;
    generator::AbstractGenerator,
    model::AbstractFittedModel=model_dict["JEM Ensemble"],
    data::CounterfactualData=counterfactual_data,
    rng::Union{Int,AbstractRNG}=Random.GLOBAL_RNG,
    _plot_title::Bool=true,
    show_factual::Bool=false,
    kwargs...,
)
    Random.seed!(rng)

@@

    _x = CounterfactualExplanations.counterfactual(ce)
    plt = Plots.plot(
        convert2image(MNIST, reshape(_x,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    if show_factual
        plt_factual = Plots.plot(
            convert2image(MNIST, reshape(x,28,28)),
            axis=nothing, 
            size=(img_height, img_height),
            title="Factual"
        )
        plt = Plots.plot(plt_factual, plt; size=(img_height*2,img_height), layout=(1,2))
    end

@@

```{julia}
_regen_all_digits = false
if _regen_all_digits
    function plot_all_digits(rng=123;verbose=true,kwargs...)
        plts = []
        for i in 0:9
            for j in 0:9
                @info "Generating counterfactual for $(i) -> $(j)"
                plt = plot_mnist(i,j;kwargs...,rng=rng)
                !verbose || display(plt)
                plts = [plts..., plt]
            end
        end
        plt = Plots.plot(plts...; size=(img_height*10,img_height*10), layout=(10,10))
        return plt
    end
    plt = plot_all_digits(generator=eccco_generator)
    savefig(plt, joinpath(output_images_path, "mnist_eccco_all_digits.png"))
end
```

@@

# Benchmark generators:
generator_dict = Dict(
    "Wachter" => WachterGenerator(λ=Λ[1]),
    "REVISE" => REVISEGenerator(λ=Λ[1]),
    "Schut" => greedy_generator,
    "ECCCo" => eccco_generator,
)
```

>>>>

```{julia}
function plot_mnist(
    factual::Int,target::Int;
    generator::AbstractGenerator,
    model::AbstractFittedModel=model_dict["JEM Ensemble"],
    data::CounterfactualData=counterfactual_data,
    rng::Union{Int,AbstractRNG}=Random.GLOBAL_RNG,
    _plot_title::Bool=true,
    show_factual::Bool=false,
    img_height::Int=180,
    kwargs...,
)
    Random.seed!(rng)

@@

    _x = CounterfactualExplanations.counterfactual(ce)
    plt = Plots.plot(
        convert2image(MNIST, reshape(_x,28,28)),
        axis=([], false),
        size=(img_height, img_height),
        title=_title
    )
    if show_factual
        plt_factual = Plots.plot(
            convert2image(MNIST, reshape(x,28,28)),
            axis=([], false),
            size=(img_height, img_height),
            title="Factual"
        )
        plt = Plots.plot(plt_factual, plt; size=(img_height*2,img_height), layout=(1,2))
    end

@@

```{julia}
_regen_all_digits = true
if _regen_all_digits
    function plot_all_digits(rng=123;verbose=true,img_height=180,kwargs...)
        plts = []
        for i in 0:9
            for j in 0:9
                @info "Generating counterfactual for $(i) -> $(j)"
                plt = plot_mnist(i,j;kwargs...,rng=rng, img_height=img_height)
                !verbose || display(plt)
                plts = [plts..., plt]
            end
        end
        plt = Plots.plot(plts...; size=(img_height*10,img_height*10), layout=(10,10), dpi=300)
        return plt
    end
    plt = plot_all_digits(generator=eccco_generator)
    savefig(plt, joinpath(output_images_path, "mnist_eccco_all_digits.png"))
end
```

@@

# Benchmark generators:
generator_dict = Dict(
    "Wachter" => WachterGenerator(λ=Λ[1], opt=eccco_generator.opt),
    "REVISE" => REVISEGenerator(λ=Λ[1], opt=eccco_generator.opt),
    "Schut" => greedy_generator,
    "ECCCo" => eccco_generator,
)
```

```{julia}
generator_params = DataFrame(
    Dict(
        :λ1 => Λ[1],
        :λ2 => Λ[2],
        :λ3 => Λ[3],
        :opt => string(typeof(eccco_generator.opt)),
        :eta => eccco_generator.opt.eta,
        :dataname => "MNIST",
    )
)
CSV.write(joinpath(params_path, "generator/mnist.csv"), generator_params)
```

### 2023-05-17 notebooks/tables.Rmd
ufff

```{r}
tab_full <- dcast(tab_valid, dataname + model + generator ~ variable)
col_names <- c(
  "Model",
  "Data",
  "Generator",
  "Cost ↓", 
  "Unfaithfulness ↓", 
  "Implausibility ↓", 
  "Redundancy ↑",
  "Uncertainty ↓",
  "Validity ↑"
)
algin_cols <- c(rep('l',3),rep('c',ncol(tab_full)-3))
kbl(
  tab_full, caption = "All results for all datasets: sample averages +/- one standard deviation over all valid counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-full} \\newline", 
  align = "c", col.names=col_names, booktabs = T, escape=F, 
  format="latex"
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  collapse_rows(columns = 1:3, latex_hline = "custom", valign = "top", custom_latex_hline = 1:2) %>%
  save_kable("paper/contents/table_all_valid.tex")
```

>>>>

```{r}
tab_full <- dcast(tab_valid, dataname + model + generator ~ variable)
col_names <- c(
  "Model",
  "Data",
  "Generator",
  "Cost ↓", 
  "Unfaithfulness ↓", 
  "Implausibility ↓", 
  "Redundancy ↑",
  "Uncertainty ↓",
  "Validity ↑"
)
algin_cols <- c(rep('l',3),rep('c',ncol(tab_full)-3))
kbl(
  tab_full, caption = "All results for all datasets: sample averages +/- one standard deviation over all valid counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-full-valid} \\newline", 
  align = "c", col.names=col_names, booktabs = T, escape=F, 
  format="latex"
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  collapse_rows(columns = 1:3, latex_hline = "custom", valign = "top", custom_latex_hline = 1:2) %>%
  save_kable("paper/contents/table_all_valid.tex")
```

## EBM

```{r}
files <- list.files("artifacts/params/")
dt <- lapply(files, function(x) {
    fread(file.path("artifacts/params/", x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
setcolorder(
  dt, 
  c(
    "dataname", "n_obs",
    "n_hidden", "n_layers", "activation", "n_ens",
    "epochs", "batch_size",
    "jem_sampling_steps", "sgld_batch_size", "lambda"
  )
)
dt_ebm <- dt[,.(dataname, jem_sampling_steps, sgld_batch_size, lambda)]
col_names <- c(
  "Dataset",
  "SGLD Steps", "Batch Size", "$\\lambda$"
)
kbl(
  dt_ebm, caption = "EBM hyperparemeter choices for our experiments. \\label{tab:ebmparams} \\newline", 
  align = "r", col.names=col_names, booktabs = T, escape=F, 
  format="latex"
) %>%
  kable_styling(font_size = 8) %>%
  kable_paper(full_width = F) %>%
  save_kable("paper/contents/table_ebm_params.tex")
```

## Experimental setup

```{r}
dt_exp <- dt[,.(dataname, n_obs, n_hidden, n_layers, activation, n_ens, epochs, batch_size)]
col_names <- c(
  "Dataset", "Sample Size",
  "Hidden Units", "Hidden Layers", "Activation", "Ensemble Size",
  "Epochs", "Batch Size"
)
header <- c(" " = 2, "Network Architecture" = 4, "Training" = 2)
kbl(
  dt_exp, caption = "Paremeter choices for our experiments. \\label{tab:params} \\newline", 
  align = "r", col.names=col_names, booktabs = T, escape=F, 
  format="latex"
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(header) %>%
  save_kable("paper/contents/table_params.tex")
```

### 2023-05-17 paper/paper.tex
ufff

Consistent with~\citet{grathwohl2020your}, we have specified $\epsilon=2$ and $\sigma=0.01$ as the default values for all of our experiments. The number of total SGLD steps $J$ varies by dataset. Following best practices, we initialize $\mathbf{x}_0$ randomly in 5\% of all cases and sample from a buffer in all other cases. The buffer itself is randomly initialised and gradually grows to a maximum of 10,000 samples during training as $\hat{\mathbf{x}}_{J}$ is stored in each epoch~\citep{du2020implicit,grathwohl2020your}.

@@

where $L_{\text{reg}}(\theta)$ is a Ridge penalty (L2 norm) that regularises energy magnitudes for both observed and generated samples~\citep{du2020implicit}. We have used varying degrees of regularization depending on the dataset.

Contrary to existing work, we have not typically used the entire minibatch of training data for the generative loss component but found that using a subset of the minibatch was often sufficient in attaining decent generative performance. This has helped to reduce the computational burden for our models, which should make it easier for others to reproduce our findings. Figures~\ref{fig:mnist-gen} and~\ref{fig:moons-gen} show generated samples for our \textit{MNIST} and \textit{Moons} data, to provide a sense of their generative property.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{../artifacts/results/images/moons_generated_JEM.png}
  \caption{Conditionally generated samples (stars) for our \textit{Moons} data using a JEM.}\label{fig:moons-gen}
\end{figure}

\subsubsection{Inference: Quantifying Models' Generative Property}

@@

\subsection{Results}\label{app:results}

>>>>

Consistent with~\citet{grathwohl2020your}, we have specified $\epsilon=2$ and $\sigma=0.01$ as the default values for all of our experiments. The number of total SGLD steps $J$ varies by dataset (Table~\ref{tab:ebmparams}). Following best practices, we initialize $\mathbf{x}_0$ randomly in 5\% of all cases and sample from a buffer in all other cases. The buffer itself is randomly initialised and gradually grows to a maximum of 10,000 samples during training as $\hat{\mathbf{x}}_{J}$ is stored in each epoch~\citep{du2020implicit,grathwohl2020your}.

@@

where $L_{\text{reg}}(\theta)$ is a Ridge penalty (L2 norm) that regularises energy magnitudes for both observed and generated samples~\citep{du2020implicit}. We have used varying degrees of regularization depending on the dataset ($\lambda$ in Table~\ref{tab:ebmparams}).

Contrary to existing work, we have not typically used the entire minibatch of training data for the generative loss component but found that using a subset of the minibatch was often sufficient in attaining decent generative performance (Table~\ref{tab:ebmparams}). This has helped to reduce the computational burden for our models, which should make it easier for others to reproduce our findings. Figures~\ref{fig:mnist-gen} and~\ref{fig:moons-gen} show generated samples for our \textit{MNIST} and \textit{Moons} data, to provide a sense of their generative property.

\import{contents/}{table_ebm_params.tex}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{../artifacts/results/images/moons_generated_JEM.png}
  \caption{Conditionally generated samples (stars) for our \textit{Moons} data using a JEM.}\label{fig:moons-gen}
\end{figure}
\subsubsection{Inference: Quantifying Models' Generative Property}

@@

Table~\ref{tab:params} provides an overview of all parameters related to our experiments. The \textit{GMSC} data were randomly undersampled for balancing purposes and all features were standardized. \textit{MNIST} data was also randomly undersampled for reasons outlined below. Pixel values were preprocessed to fall in the range of $[-1,1]$ and a small Gaussian noise component ($\sigma=0.03$) was added to training samples following common practices in the EBM literature.

\subsubsection{Compute}

To enable others to easily replicate our experiments, we have chosen to work with small neural network architectures and randomly undersampled large datasets where appropriate. All of our experiments could then be run locally on a personal machine. The longest runtimes we experienced for model training and counterfactual benchmarking were on the order of 8-12 hours (\textit{MNIST} data). For the synthetic data, all experiments could be completed in less than an hour.

We have summarised the system information below:

\textbf{Software}:

\begin{itemize}
  \item System Version: macOS 13.3.1
  \item Kernel Version: Darwin 22.4.0
  \item Boot Volume: Macintosh HD
  \item Boot Mode: Normal
\end{itemize}

\textbf{Hardware}:

\begin{itemize}
  \item Model Name: MacBook Pro
  \item Model Identifier: MacBookPro16,1
  \item Processor Name: 8-Core Intel Core i9
  \item Processor Speed: 2.3 GHz
  \item Number of Processors: 1
  \item Total Number of Cores: 8
  \item L2 Cache (per Core): 256 KB
  \item L3 Cache: 16 MB
  \item Hyper-Threading Technology: Enabled
  \item Memory: 32 GB
\end{itemize}

\import{contents/}{table_params.tex}
\subsection{Results}\label{app:results}

Figure~\ref{fig:mnist-eccco} shows examples of counterfactuals for \textit{MNIST} data where the underlying model is our \textit{JEM Ensemble}. Original images are shown on the diagonal and the corresponding counterfactuals are plotted across rows.

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{../artifacts/results/images/mnist_eccco_all_digits.png}
  \caption{Counterfactuals for \textit{MNIST} data and our \textit{JEM Ensemble}. Original images are shown on the diagonal with the corresponding counterfactuals plotted across rows.}\label{fig:mnist-eccco}
\end{figure}

Table~\ref{tab:results-full} reports all of the evaluation metrics we have computed. Table~\ref{tab:results-full} reports the same metric for the subset of valid counterfactuals. The `Unfaithfulness' and `Implausibility' metrics have been discussed extensively in the body of the paper. The `Cost' metric relates to the distance between the factual and the counterfactual. The `Redundancy' metric measures sparsity as the percentage of features that remain unperturbed (higher is better). The `Uncertainty' metric is just the average value of the smooth set size penalty (Equation~\ref{eq:setsize}). Finally, `Validity' is the percentage of valid counterfactuals.
